---
title: "Making Predictions About Brain Weight"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

The objective of this project is to analyze data related to brain weight and head size and how it differs between male versus female, and adult versus juvenile humans.

## Background

The data for this project was found on Kaggle. Techniques learned in BIOL-3100 will be used to clean and analyze the data, as well as make models and predictions.

## Data
#loading libraries
```{r}
library(tidyverse)
library(GGally)
library(ggplot2)
library(MASS)
library(caret)
library(easystats)
```
```{r}
brain_dat <- read.csv('FinalProject_Dataset.csv.csv')

categorized_dat <- brain_dat %>% 
  mutate(Gender = case_when(Gender == 1 ~ "Male",
                            Gender == 2 ~ "Female"),
         Age.Range = case_when(Age.Range == 1 ~ "Minor",
                               Age.Range == 2 ~ "Adult"))
```
#Visualizing Relationships
```{r}
ggpairs(categorized_dat)
#apparent linear trend between head size and brain weight; normal distribution.

str(categorized_dat)
dim(categorized_dat)#237 4

#Visualizing the relationship between head size and brain weight.
categorized_dat %>% 
  ggplot(aes(x = Brain.Weight.grams., y = Head.Size.cm.3.))+
  geom_point()+
  geom_smooth(method = 'lm', se = F)+
  theme_minimal()+
  labs(title = 'Head Size Versus Brain Weight with Trend Line',
       x = 'Brain Weight (g)',
       y = 'Head Size (cm^3)')

#Visualizing the distributions
brain_dat %>% 
  mutate(Gender = as.factor(Gender)) %>% 
  ggplot(aes(x = Head.Size.cm.3., fill = Gender)) +
  geom_histogram(color = "black", position = "dodge") +
  labs(
    title = "Head Size Distribution by Sex",
    x = "Head Size (cm^3)",
    y = "Frequency",
    fill = "Sex"
  ) +
  geom_vline(aes(xintercept = mean(Head.Size.cm.3.)), color = "red", linetype = "dashed") +
  theme_light() +
  scale_fill_manual(values = c("skyblue", "darkblue"),
                    labels = c("Male", "Female")) +  # Custom color for fill
  theme(legend.position = "right")  # Position legend at the top

brain_dat %>% 
  mutate(Gender = as.factor(Gender)) %>% 
  ggplot(aes(x = Brain.Weight.grams., fill = Gender)) +
  geom_histogram(color = "black", position = "dodge") +
  labs(
    title = "Brain Weight Distribution by Sex",
    x = "Brain Weight (g)",
    y = "Frequency",
    fill = "Sex"
  ) +
  geom_vline(aes(xintercept = mean(Brain.Weight.grams.)), color = "red", linetype = "dashed") +
  theme_light() +
  scale_fill_manual(values = c("orange", "maroon"),
                    labels = c("Male", "Female")) +  # Custom color for fill
  theme(legend.position = "right")  # Position legend at the top
```
#Finding the best model
```{r}
full_model <- glm(data = categorized_dat,
                  formula = Brain.Weight.grams. ~ Gender*Age.Range*Head.Size.cm.3.)#do all the possibilities in a full model.

summary(full_model)
stepwise_mod <- stepAIC(full_model, direction = 'both')#this function checks all possibilities and chooses the best model.
stepwise_mod$formula
#Brain.Weight.grams. ~ Gender + Age.Range + Head.Size.cm.3. + 
#Gender:Age.Range + Age.Range:Head.Size.cm.3.

best_model <- glm(data = categorized_dat,
                  formula = stepwise_mod$formula)

summary(best_model)
predict(best_model)


categorized_dat$Predictions <- predict(best_model)
```
#Training the model
```{r}
summary(categorized_dat$Head.Size.cm.3.)
id <- createDataPartition(categorized_dat$Head.Size.cm.3., list = F)
dat_train <- categorized_dat[id, ]
dim(dat_train)#119 5
dim(categorized_dat)#237 5

dat_test <- categorized_dat[-id, ]

train_mod <- glm(data = dat_train,
                 formula = stepwise_mod$formula)
dat_test$pred <- predict(train_mod, dat_test)

actual <- dat_test$Brain.Weight.grams.
predicted <- dat_test$pred

dat_test$residuals <- actual - predicted

ggplot(dat_test, aes(x = pred, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted",
       x = "Predicted Brain Weight",
       y = "Residuals") +
  theme_minimal()
#The residuals plot looks pretty good. The residuals are centered around zero and don't fan out or curve, so there's no big red flag. It suggests the linear model is reasonably appropriate for the data.

dat_test %>% 
  ggplot(aes(x = Brain.Weight.grams., y = pred))+
  geom_point(alpha = 0.6)+
  geom_smooth(method = 'lm', se = F)+
  labs(title = 'Prediction Accuracy',
       x = 'Actual Brain Weight',
       y = 'Predicted Brain Weight')+
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+#abline shows what we'd want if the model were perfect.
  theme_minimal()
```

```{r}
categorized_dat %>% 
  ggplot(aes(x = Head.Size.cm.3., y = Predictions, color = Gender)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Predicted Brain Weight by Sex",
       x = "Head Size (cm^3)",
       y = "Predicted Brain Weight",
       color = "Sex")+
  theme_minimal()
```

```{r}
categorized_dat %>% 
  ggplot(aes(x = Head.Size.cm.3., y = Predictions, color = Age.Range)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Predicted Brain Weight by Age",
       x = "Head Size (cm^3)",
       y = "Predicted Brain Weight",
       color = "Age")+
  theme_minimal()
```
#Plotting the model with the real data
```{r}
categorized_dat %>% 
  ggplot(aes(x = Brain.Weight.grams., y = Head.Size.cm.3.))+
  geom_point()+
  geom_point(aes(x = Predictions, y = Head.Size.cm.3., color = 'red'))+
  geom_smooth(method = 'lm', se = F)+
  theme_minimal()+
  labs(title = 'Head Size Versus Brain Weight with Predictions',
       x = 'Brain Weight (g)',
       y = 'Head Size (cm^3)')+
  theme(legend.position = "none")
```

```{r}
mse <- mean((categorized_dat$Brain.Weight.grams. - categorized_dat$Predictions)^2)
print(mse)#mse
```




